Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2023-12-19 paper.tex
Paper organization

In this contribution we continue our initiative to develop a simple, efficient
and open-source solution to compute all classical correlation functions on CPU
and GPU architectures -- \code{CorrelationFunctions.jl} \cite{CFsjlpaper}. Here
we build upon previous results in computations of higher order functions with n
> 2, most notably by Malmir et al. \cite{malmir2018}, and our aim is to develop
robust and computationally efficient methods. In addition, we also considered
the case of anisotropic structures and how to describe them with 3-point
correlation functions similar to directional 2-point CFs \cite
{10.1063/1.4867611,EPL1}.

\highlight{The paper is organized as follows:}

\section{Methods}
\subsection{3-point correlation functions}
Let $A_1, A_2, \dots, A_n$ be pairwise disjoint subsets of a set
$A \subset \mathbb{R}^n$, where the index can be interpreted as a phase on the
image $A$. We introduce a family of indicator functions
$I_n(x) : A \rightarrow \{0,1\}$ defined as:
\begin{equation}
  I_n(x) = \left\{
  \begin{array}{ll}
    1 & \quad x \in A_n \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

@@

\section{Results}
\subsection{Verification}
To verify correctness of computation of $S_3$ and $C_3$ functions we use the
following simple relations:
\begin{align}
  S_3^n (x, 0) = S_3^n (0, x) &= S_2^n(x) \\
  C_3^n (x, 0) = C_3^n (0, x) &= C_2^n(x) \\
  \lim_{\substack{x_1 \to \infty \\ x_2 \to \infty}} S_3^n (x_1, x_2) &= \phi_n^3
\end{align}
Here $\phi_n$ is a fraction of phase $n$ in $A$.

@@

\section{Discussion}
In this paper we have provided two approaches to compute three-point correlation
function:
\begin{enumerate}
\item Computation in frequency domain using Fourier transform. This approach has
  computational complexity $O(n^2 \log n)$ and computes the whole correlation
  map. Unfortunately, it requires $O(n^2)$ additional memory and therefore can
  be used only with small inputs. Moreover, the whole Fourier image of the input
  is needed even if we need only a small part of the correlation map.
\item Computation in spatial domain. This approach is implemented by us in
  CorrelationFunctions.jl package. This approach calculates the full map with
  complexity $O(n^3)$ but uses only $O(n)$ additional memory for shifted
  versions of the input. For our purposes the whole map is not needed and
  therefore the amount of calculations can be significantly reduced.
\end{enumerate}
It's tempting to create an algorithm which unites advantages of both approaches
(fast speed of the first and low memory consumption of the second) which
computes some desired slices of correlation map. It is known that Fourier
transform is the only linear transform which allows computation of convolution
in linear time. It is still an open question if there are linear transforms which
allow computation of convolution faster than a naïve approach
\cite{stone2008uniqueness,stone1998convolution}.

>>>>

In this contribution we continue our initiative to develop a simple, efficient
and open-source solution to compute all classical correlation functions on CPU
and GPU architectures -- \code{CorrelationFunctions.jl} \cite{CFsjlpaper}. Here
we build upon previous results in computations of higher order functions with
$n > 2$, most notably by Malmir et al. \cite{malmir2018}, and our aim is to
develop robust and computationally efficient methods. In addition, we also
considered the case of anisotropic structures and how to describe them with
3-point correlation functions similar to directional 2-point CFs
\cite{10.1063/1.4867611,EPL1}.

The paper is organized as follows: in \cref{sec:methods} we provide
mathematical background and describe methods for computation of 3-point
correlation functions. Verification of these methods along with some results for
real-world samples (limestone, soil etc.) is described in
\cref{sec:results}. Section \ref{sec:discussion} provides the discussion, a
comparison with other existing methods and summarizes possible directions for
improvement.

\section{Methods}
\label{sec:methods}
\subsection{3-point correlation functions}
Let $A_1, A_2, \dots, A_n$ be pairwise disjoint subsets of a set
$A \subset \mathbb{R}^n$, where the index can be interpreted as a phase on the
image $A$. We introduce a family of indicator functions
$I_n(x) : A \rightarrow \{0,1\}$ defined as:
\begin{equation}
  I_n(x) = \left\{
  \begin{array}{ll}
    1 & \quad x \in A_n \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

@@

\section{Results}
\label{sec:results}
\subsection{Verification}
To verify correctness of computation of $S_3$ and $C_3$ functions we use the
following simple relations:
\begin{align}
  S_3^n (x, 0) = S_3^n (0, x) &= S_2^n(x) \\
  C_3^n (x, 0) = C_3^n (0, x) &= C_2^n(x) \\
  \lim_{\substack{x_1 \to \infty \\ x_2 \to \infty}} S_3^n (x_1, x_2) &= \phi_n^3
\end{align}
Here $\phi_n$ is a fraction of phase $n$ in $A$.

@@

\section{Discussion}
\label{sec:discussion}
In this paper we have provided two approaches to compute three-point correlation
function:
\begin{enumerate}
\item Computation in frequency domain using Fourier transform. This approach has
  computational complexity $O(n^2 \log n)$ and computes the whole correlation
  map. Unfortunately, it requires $O(n^2)$ additional memory and therefore can
  be used only with small inputs. Moreover, the whole Fourier image of the input
  is needed even if we need only a small part of the correlation map.
\item Computation in spatial domain. This approach is implemented by us in
  CorrelationFunctions.jl package. This approach calculates the full map with
  complexity $O(n^3)$ but uses only $O(n)$ additional memory for shifted
  versions of the input. For our purposes the whole map is not needed and
  therefore the amount of calculations can be significantly reduced.
\end{enumerate}
It's tempting to create an algorithm which unites advantages of both approaches
(fast speed of the first and low memory consumption of the second) which
computes some desired slices of correlation map. It is known that Fourier
transform is the only linear transform which allows computation of convolution
in linear time. It is still an open question if there are linear transforms which
allow computation of convolution faster than a naïve approach
\cite{stone2008uniqueness,stone1998convolution}.
